Accelerated Inference for Partially Observed Markov Processes using Automatic Differentiation

### 2024-06-14 paper/ms.tex
update from overleaf

We start by presenting a new construction of the gradient estimator of \cite{scibior21} and \cite{poyiadjis11}.
Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a {\it Measurement Off-Parameter} (MOP) particle filter.
We proceed to generalize the MOP particle filter by adding a discounting factor, $\alpha \in [0,1]$, which interpolates between the  biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$.
The bias-variance tradeoff induced by $\alpha$ suggests, both in theory and in practice, the use of $\alpha$ values strictly between 0 and 1.

@@

Directly differentiating through a particle filter, for a fixed value of $\omega$, leads to a biased gradient estimator due to the discreteness of the resampling operation \cite{corenflos21}.
We avoid this issue by constructing a particle filtering algorithm called {\it Measurement Off-Parameter with discount factor} $\alpha$ (MOP-$\alpha$), defined by the pseudocode in Algorithm~\ref{alg:mop}.
In short, instead of differentiating through a Monte Carlo algorithm directly, we instead use AD to differentiate through a series of measurement density ratios and a differentiable simulator.
Note that the plug-and-play property forbids access to transition densities but permits access to the density of the measurements conditional on the value of the latent process.
The simulator can be differentiated using AD as long as the underlying computer code for the simulator is differentiable, and so MOP-$\alpha$ is applicable to a broad class of POMP models.

The MOP-$\alpha$ algorithm resamples the particles according to an arbitrary resampling rule that depends on both the target parameter value and a baseline parameter value, and this explains the name {\it measurement off-parameter}.
Similar quantities are called {\it off-policy} in reinforcement learning, and a parameter in a POMP model is analogous to a policy in a corresponding control model \ed{Kevin, would it be good to add your favorite RL reference here? Also, you may want to edit this sentence. I think something like this will help the reader to ``get'' our terminology.}.
Specifically, MOP-$\alpha$ evaluates the likelihood at some $\theta \in \Theta$, but instead resamples the particles according to the indices generated by a vanilla particle filter run at a {\it baseline parameter}, $\phi \in \Theta$.
This ensures that the resampling indices are invariant to $\theta$ when $\omega$ and $\phi$ are fixed, bypassing the issue of Monte Carlo resampling.
We show in Theorem~\ref{thm:mop-targeting} that MOP-$\alpha$ exactly targets the filtering distribution when $\theta=\phi$ or $\alpha=1$.

@@

\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations, performing IF2 alone ultimately fails to achieve the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter}). Conversely, when we tried gradient descent with MOP alone, we encountered many failed searches. This is a difficult, nonconvex, and noisy problem, \ed{SAY HOW MANY PARAMETERS ARE ESTIMATED HERE} and the search gets stuck in local minima and saddle points, failing to approach the MLE.

@@

\ed{Labeling A and B would be more conventional than Left and Right, and also easier to refer to in the text.}

@@

\begin{figure}[t!]
    \centering
    \includegraphics[width=\textwidth/\real{1.25}]{../imgs/pmcmc/nuts_eb.png}
    \caption{Convergence diagnostics for NUTS with the informative nonparametric empirical Bayes prior from applying KDE to preliminary searches using IF2. We display the results for the trend parameter in the Dhaka cholera model of \cite{king08}. The NUTS sampler mixes quickly, and the posterior estimates from each chain share roughly the same posterior mode.}
    \label{fig:nuts-eb}
\end{figure}

@@

We also show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( NpG'(\theta)^2\big/J \big)$, and note that \cite{poyiadjis11} establish that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p, G'$, etc. \ed{complete the etc to be more precise}. 
Combining that result with Theorem \ref{thm:mop-biasvar} shows that MOP-$\alpha$ interpolates between MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$.

@@

MOP-$\alpha$ and IFAD are fast algorithms, both in theory and practice. 
In line with the cheap gradient principle of \cite{kakade2019provably}, getting a gradient estimate from MOP-$\alpha$ takes no more than 6 times that of the runtime of the particle filter (in practice, the multiplier is 3.75). 
MOP-$\alpha$ and IFAD therefore share the same $O(NJ)$ time complexity as the particle filter, unlike the $O(NJ^2)$ complexity of \cite{corenflos21} and Algorithm 2 in \cite{poyiadjis11, scibior21}.

@@

\arxiv{
\section*{Acknowledgments}
This research was supported by National Science Foundation grant DMS-1761603. 
We acknowledge communications with Nicolas Chopin. 
}{
\showmatmethods{} % Display the Materials and Methods section
\acknow{Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.}
\showacknow{} % Display the acknowledgments section
}

>>>>

We start by presenting a new construction of the gradient estimator of \cite{scibior21} and \cite{poyiadjis11}.
Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a {\it Measurement Off-Parameter} (MOP) particle filter. In short, \textit{instead of directly differentiating through a basic particle filter, MOP uses AD to differentiate through a series of measurement density ratios and a differentiable simulator}.
We proceed to generalize the MOP particle filter by adding a discounting factor, $\alpha \in [0,1]$, which interpolates between the  biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$.
The bias-variance tradeoff induced by $\alpha$ suggests, both in theory and in practice, the use of $\alpha$ values strictly between 0 and 1.

@@

Directly differentiating through a particle filter, for a fixed value of $\omega$, leads to a biased gradient estimator due to the discreteness of the resampling operation \cite{corenflos21}.
We avoid this issue by constructing a particle filtering algorithm called {\it Measurement Off-Parameter with discount factor} $\alpha$ (MOP-$\alpha$), defined by the pseudocode in Algorithm~\ref{alg:mop}.
In short, \textit{instead of directly differentiating through a basic particle filter, we instead use AD to differentiate through a series of measurement density ratios and a differentiable simulator}.
Note that the plug-and-play property forbids access to transition densities but permits access to the density of the measurements conditional on the value of the latent process.
The simulator can be differentiated using AD as long as the underlying computer code for the simulator is differentiable, and so MOP-$\alpha$ is applicable to a broad class of POMP models.

The MOP-$\alpha$ algorithm resamples the particles according to an arbitrary resampling rule that depends on both the target parameter value and a baseline parameter value, and this explains the name {\it measurement off-parameter}.
This is intended as an analogy to {\it off-policy} learning in reinforcement learning.
Specifically, MOP-$\alpha$ evaluates the likelihood at some $\theta \in \Theta$, but instead resamples the particles according to the indices generated by a vanilla particle filter run at a {\it baseline parameter}, $\phi \in \Theta$.
This ensures that the resampling indices are invariant to $\theta$ when $\omega$ and $\phi$ are fixed, bypassing the issue of Monte Carlo resampling.
We show in Theorem~\ref{thm:mop-targeting} that MOP-$\alpha$ exactly targets the filtering distribution when $\theta=\phi$ or $\alpha=1$.

@@

\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations, performing IF2 alone ultimately fails to achieve the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter}). Conversely, when we tried gradient descent with MOP alone, we encountered many failed searches. This is a difficult, nonconvex, and noisy problem with $18$ parameters, and the search gets stuck in local minima and saddle points, failing to approach the MLE.

@@

%% \ed{Labeling A and B would be more conventional than Left and Right, and also easier to refer to in the text.}

@@

\begin{figure}[t!]
    \centering
    \includegraphics[width=\textwidth/\real{1.25}]{../imgs/pmcmc/nuts_eb.png}
    \caption{Convergence diagnostics for NUTS with the informative nonparametric empirical Bayes prior from applying KDE to preliminary searches using IF2, with $4$ chains. We display the results for the trend parameter in the Dhaka cholera model of \cite{king08}. The NUTS sampler mixes quickly, and the posterior estimates from each chain share roughly the same posterior mode.}
    \label{fig:nuts-eb}
\end{figure}

@@

We also show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( NpG'(\theta)^2\big/J \big)$, and note that \cite{poyiadjis11} establish that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p$ and $G'$. 
Combining that result with Theorem \ref{thm:mop-biasvar} shows that MOP-$\alpha$ interpolates between MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$. The phase transition arises as even though the particle filter has forgetting properties \cite{karjalainen23}, the resulting derivative estimate of \cite{poyiadjis11} does not, and we require both for the variance reduction.

@@

MOP-$\alpha$ and IFAD are fast algorithms, both in theory and practice. 
In line with the cheap gradient principle of \cite{kakade2019provably}, getting a gradient estimate from MOP-$\alpha$ takes no more than 6 times that of the runtime of the particle filter. In our simulations, MOP-$\alpha$ took 3.75 times the runtime of the particle filter. 
MOP-$\alpha$ and IFAD therefore share the same $O(NJ)$ time complexity as the particle filter, unlike the $O(NJ^2)$ complexity of \cite{corenflos21} and Algorithm 2 in \cite{poyiadjis11, scibior21}.

@@

\arxiv{
\section*{Acknowledgments}
This research was supported by National Science Foundation grant DMS-1761603. 
We thank Nicolas Chopin for helpful communications regarding the strong law of large numbers for off-parameter resampled particle filters. 
}{
\showmatmethods{} % Display the Materials and Methods section
\acknow{Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.}
\showacknow{} % Display the acknowledgments section
}
